---
title: "Post Lab 1 Practice"
author: "Nathaniel Grimes"
format: 
  html:
    code-fold: true
    embed-resources: true
execute:
  eval: true
  echo: true
  warning: false
  message: false
---

# Post-Lab Practice

Create a new repository in your GitHub account. Clone this into R Studio (no need to fork, it's already in your account).

In this new repository, create a new Quarto document and set it up:

-   Code folding (either `true` (fold, but hide it) or `show` (fold, but show it))
-   Echo code to the document, but suppress warnings and messages

Delete the template text, and create a new code chunk. Attach the `tidyverse` package. Attach the `gapminder` package (install it if you need to!).
```{r}
library(tidyverse)
library(gapminder)
```

## Wrangling Practice

1. Examine the `relig_income` data. Is it tidy? Which of these functions would you use to help you make it tidy?

  a. `mutate()`
  
  b. `separate()`
  
  c. `pivot_longer()`
  
  d. `select()`
  
2. Use the function you selected in question 1 to make `relig_income` tidy

3. What does the following code do?

```{r}
#| eval: false

lizard_clean<-lizard %>% 
  select(spp,sex,SV_length,weight) %>% 
  filter(sex!="J") %>% 
  drop_na() %>% 
  group_by(spp,sex) %>% 
  filter(n()>=6)
```



## Analysis Work:

Use the `gapminder` life expectancy data from `gapminder` package (install if needed).
```{r}
# Load data
df <- gapminder
head(df)
```

Conduct an analysis on whether life expectancy is increasing over time. 
```{r}
# visualize life expectance over time
df %>% 
  ggplot(aes(x=year, y=lifeExp, color=continent)) +
  geom_point(alpha=0.01)+
  geom_smooth()+
  theme_minimal()+
  labs(
    title="Life Expectancy Over time",
    x='year',
    y='Life Expectancy'
    
  )
```


Run a linear regression on the data with `gdpPercap` and `continent` as covariates.
```{r}
model1 <- lm(lifeExp ~ gdpPercap + continent, data = df)
summary(model1)

model2 <- lm(lifeExp ~ log(gdpPercap) + continent, data=df)
summary(model2)
```


Plot the observed trends however you see fit.
```{r}
# plot GDP and life expectancy
df %>% ggplot(
  aes(x=log(gdpPercap), y = lifeExp, color = continent)) +
  geom_point()+
  theme_minimal()+
  labs(
    title='life expectancy vs GDP log',
    x='log gdp',
    y='life expectancy'
  )
```

```{r}
df %>% ggplot(
  aes(x=gdpPercap, y = lifeExp, color = continent)) +
  geom_point()+
  theme_minimal()+
  labs(
    title='life expectancy vs GDP',
    x='gdp',
    y='life expectancy'
  )
```


```

## Followup questions:

-   Write a sentence or two describing the trends you see in the plots. Which model (log or not) did you choose and why?

**log, because plot is less clusters with log(gdp)**

-   Write a sentence or two to explain the coefficients on your linear model output, including statistical significance.
**For 1 increase in log(GDP), life expectancy increases by 0.06422 years. Americas had 7 more years of LE than africa. Life e is strongly related to wealth.**
    -   Which continent is the "reference" continent?
    **the one that's now shown as coefficient: in this case Africa**
    
    -   What does the "year" coefficient represent? 
    **my model didn't have year as a predictor, but if year is included in the model, the year coefficient would represent the average increase in life expectancy per year, holding GDP and continent constant.**but it also could mean technical progress over time that benefits LE independent of economic growth. 
    
    -   What do the coefficients on each continent represent? Do these make intuitive sense?
    **how much would model shift from reference continent**
    
-   Does a linear model seem like a good way to approach these data? Why or why not?
  **not sure if a linear model capture demographic transition, where diminishing return is captured**
  

Render your Quarto document, and then stage, commit, and push back to your Github repository. Make sure the updates are reflected online!





